---
title: "DATAX500"
author: "1631819"
date: "2026-03-22"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#tinytex::install_tinytex(force = TRUE)
```
--------------------------------------------------------------------------------
[Github Repository](https://github.com/BrewBarred/DATAX500_A1)

[Moodle Submission](https://elearn.waikato.ac.nz/mod/assign/view.php?id=2241094&action=editsubmission)

[Download Handout](https://elearn.waikato.ac.nz/pluginfile.php/4467078/mod_assign/introattachment/0/Assignment%201.pdf?forcedownload=1)
--------------------------------------------------------------------------------

# Assignment 1

1. Using the '_heart.csv_' data-set, calculate the following statistical measures for age in _R_ 
---------------------------------------------------------------------------------------------------------------------
   
First, we load the data-set:
   
  ```{r load, echo=TRUE}
  # load automatically via git files
  heart<-read.csv("Heart.csv")
  
  
  ## Alternatively, load the dataset manually using:
  # heart<-read.csv(file.choose())
  ```
      
(6 marks)

### a) Mean 
               
  ```{r mean, echo=TRUE}      
  # assign age column to "age" var for easier processing
  age<-heart$age
  
  # calculate the mean age
  mean_age<-mean(age)
  mean_age
  
  # round to 2 d.p.
  mean_age<-round(mean_age, 2)
  mean_age
  ```
  
(2 marks)

### b) Median 
           
  ```{r median, echo=TRUE}      
  # calculate the median age
  median_age<-median(age)
  median_age
  ```
  
(2 marks)

### c) Sample standard deviation 

  ```{r sd, echo=TRUE}      
  # calculate the sample standard deviation
  sd_age<-sd(age)
  sd_age
  
  # round sample standard deviation to 2 d.p.
  sd_age<-round(sd_age, 2)
  sd_age
  ```

(2 marks)

---------------------------------------------------------------------------------------------------------------------

2. A researcher wishes to investigate the relationship between _gender_ and _cholesterol levels_
---------------------------------------------------------------------------------------------------------------------
   
Key-points to notice here is that **we are measuring cholesterol levels** 
based on gender, making **cholesterol** the _dependent variable_, and 
**gender** the _independent variable_.

Using the heart.csv data set, select and perform an _appropriate statistical_ 
_test_ in _R_ and interpret the output. Your answer should include a **null**
**hypothesis**, an **alternative hypothesis **, and a **p-value ** 

Since _gender_ (categorical) and _cholesterol_ (continuous) are **independent** 
from each other, we can conduct an independent (two-sampled) t-test to test the 
following hypotheses:
   
   
### Null hypothesis:
        
  - There will be no difference in mean cholesterol levels between genders.
    (H0: U1 = U2)
   
### Alternative hypothesis:
        
  - There will be a significant difference in mean cholesterol levels between 
    genders.
    (H1: U1 != U2)
        
### Results:
      
  ```{r p-value, echo=TRUE}
  # peek at the dataset so we can view the column names
  head(heart, n=2)
  
  # perform a t-test on the heart data set between chol and gender
  t.test(chol~gender, data=heart)
  ```

### _p_-value:

  _p_ = 0.003 (3.d.p.)

### **Interpretation:**
  
  - The Welch two-sample t-test provides **evidence to reject the null 
    hypothesis**, that **there will be no difference in mean cholesterol 
    levels between genders, _p_=0.003** (3 d.p.).
  
  - This interpretation is further supported with **a large t-value of 
    2.98.**
  
  - **Further testing is recommended** to determine the source or 
    magnitude of this cholesterol difference.
    
  - With _95% confidence_, we can assume that the **true mean difference**
    will be somewhere between 7.29 and 36.17 (2 d.p.)**.
    
  - Broadly analyzing the data, an apparent **mean difference of ~22** between 
    group 0 and group 1 exists, providing more **evidence to reject the null** 
    hypothesis, that there is no difference between the means, regardless of 
    gender (for each group).
      
(5 marks)


---------------------------------------------------------------------------------------------------------------------

3. A researcher wishes to investigate the predictors of maximum heart rate. Using the data set provided, fit an appropriate statistical model of your choice. Make sure to include age as one of the predictors 
---------------------------------------------------------------------------------------------------------------------
  
### a) What are the null and alternative (two-sided) hypotheses associated with age?
        
  - **Null hypothesis:** H0: B1 = 0
  
    - The regression coefficient for age is zero (i.e., age has no linear 
      effect on maximum heart rate.)
      
    _**Note:** This means that for every _x_ increase in age, no predictable change in _y_ is observed._
  
  - **Alternative hypothesis:** H1: B1 != 0
  
    - There will be a significant relationship between age and maximum
      heart-rate.

(2 marks)

### b) What is the estimated coefficient of age and what does this imply in terms of the relationship between age and maximum heart rate? 
      
  ```{r coefficient, echo=TRUE}
  # again, peek at the dataset so we can see what we're working with
  head(heart, n=2)
  
  # store the mean maxheartrate in a variable
  rate<-heart$maxheartrate
  
  # create a multiple linear regression model to predict max heart rate 
  # given an age and gender
  model<-lm(rate ~ age + gender, data=heart)
  
  # ouput the model summary for analysis
  summary(model)
  ```

### Estimated coefficient of age:
  
  -1.0277
  
### Implications (in terms of relationship with max heart rate):
  
  This model indicates a _negative linear_ relationship. For _every
  one-year increase in **age**, the **maximum heart rate** is estimated
  to **decrease by ~1.028 bpm (3 d.p.)**_, regardless of gender.
  
(3 marks)
     
### c) Comment on the predictors you have included in the model and justify the inclusion of these variables in your model 
        
I chose to use **age** and **gender** for my predictors in this model. These are 
both _common demographic predictors_ and often enable clearer, more informative 
interpretations. 

It is understood that ones cardiovascular health typically deteriorates with 
age, but it's unclear if this is true for both Men, and Women. Using **age** and 
**gender** as _predictors_ in this model will allow us to estimate the
relationship between **age** and **maximum heart rate** while **controlling for**
**gender**

  - **Age:** _Age_ is a **continuous** variable, making it **suitable for** 
             **regression**. I chose this not only due to its common use 
             in other studies, especially health studies, but also because
             an individual's maximum heart rate is likely to decline as they get
             older - this predictor allows us to quantify that change.
  
  - **Gender:** _Gender_ is a categorical variable (0=male, 1=female).
                Controlling for gender helps to distinguish differences between
                Males and Females and is likely to increase the accuracy of our
                estimations.
        
(3 marks)
     
     
### d) Interpret the model output in plain English. Your interpretation must include the p-value associated with each coefficient (slopes and intercept) test, and R-squared value 
     
The model provides the following insights:

- Age is a highly significant predictor (_p_ = 6.17e-13, _t_ = -7.542). We can 
  say with 99.9% confidence that as age increases, maximum heart rate decreases,
  regardless of gender.
  
- Gender (_p_ = 0.058 (3.d.p.)) is slightly above the the 0.05 threshold, 
  suggesting that while there is a trend, gender is not a statistically 
  significant predictor in this model, at least, not at the 95% confidence 
  level.

- Intercept (_p_ < 2e-16) is highly significant, much lower than the threshold
  of 0.05. This suggests that the baseline of this model is statistically valid.
  
- Model accuracy: The multiple R squared value of 0.1683 suggests that only
                  approximately 16.8% of the variance in our data can be
                  explained by age and gender. While some relationship exists,
                  and is significant, there are probably some other factors
                  that could better predict the change in maximum heart rate 
                  as age increases.

In summary, we have strong evidence to reject the null hypothesis for age, that 
the regression coefficient will be zero, regardless of gender.

(3 marks)

### e) Comment on whether the assumptions of a linear regression model have been met. Include appropriate statistical plots to support your answer

Appropriate statistical plots:
        
  ```{r plots, echo=TRUE}
  # generate some plots to evaluate the regression assumptions:
    ## Linearity, Independence, Homoscedasticity and Normality
  
  # create a 2x2 grid of diagnostic plots
  #par(mfrow=c(2,2)) # uncomment for small grid-style output
  
  # plot appropriate statistical plots
  plot(model)
  ```
  
### e) Commentary on Model Assumptions

Based on the diagnostic plots, the assumptions for this linear regression are largely satisfied, though there are a
few minor deviations:

### Linearity (Residuals vs Fitted): 

  The red line is relatively flat and stays close to the zero line. This 
  suggests that the relationship between the predictors (age and gender) and 
  maximum heart rate is correctly modeled as linear. There is no obvious "U" 
  shape that would suggest we missed a non-linear trend.

###Normality (Q-Q Residuals): 

  The points follow the dashed diagonal line very closely through the middle. 
  There is a tiny bit of "wavering" at the extreme ends (specifically points 
  273 and 239), but it's not severe enough to violate the assumption of normally 
  distributed residuals.

### Homoscedasticity (Scale-Location):

  The red line is nearly horizontal, and the "cloud" of points has a fairly 
  consistent vertical spread across the whole x-axis. This indicates constant 
  variance (no "fan" shape), meaning the model is equally accurate 
  (or inaccurate) across all age groups.

### Influential Outliers (Residuals vs Leverage): 

  All points fall well within the dashed red boundaries for Cook’s Distance. 
  While points 273, 239, and 85 are labeled as having higher leverage or larger 
  residuals, they aren't "pulling" the model enough to be considered problematic 
  outliers that would invalidate the results.
  
### Independence (Study design):

  Since each observation in this data set represents a separate, unrelated 
  individual patient with no repeated measures, independence can be assumed.

(3 marks)

(14 marks)     
        
---------------------------------------------------------------------------------------------------------------------
 
Data dictionary:
---------------------------------------------------------------------------------------------------------------------

cp: chest pain type

  1. typical angina
  2. atypical angina 
  3. non-anginal pain
  4. asymptomatic

trtbps:

- resting blood pressure (in mm Hg)

chol:

- cholesterol (in mg/dl)

gender: 

0. male;
1. female;

fbs: (fasting blood sugar > 120 mg/dl)

0. False
1. True

rest_ecg: resting electrocardiographic result

0. Normal
1. Having ST-T wave abnormality (T wave inversions and/or ST elevation or 
   depression of > 0.05 mV)
2. Showing probably or definite left ventricular hypertrophy by Estes' criterion

---------------------------------------------------------------------------------------------------------------------

## Regression Assumption Notes:

### Linearity: 
  - **Residual/fitted plots** should be relatively flat and close to zero.
  - A curved or "U-shaped" line suggests a non-linear relationship 
    (consider transforming or adding polynomial terms)

### Normality:
  - The **Q-Q Residuals plots** should follow the diagonal dashed line closely.
  - Minor deviation at the tails is acceptable, but strong "S-curves" or heavy
    tailing suggests non-normal residuals 
    (consider transforming the response variable)

### Homoscedasticity: 
  - The **Scale-Location plots'** red line should be nearly horizontal with an even
    spread of points. A "fan" shape (spread increasing left to right) indicates 
    heteroscedasticity (constant variances is violated, consider transformation 
    or weighted regression)

### Independence: 
  - Cannot be visually verified, but rather, justified by study design. If each
    observation is a separate, unrelated individual/event with no repeated 
    measures or time-series structure, independence can be reasonably assumed.
    
### Influential Outliers:
  - Any points outside Cook's Distance boundaries (dashed red lines) in the 
    **Residuals vs. Leverage plot** are influential, and may be distorting the
    model (consider investigating or removing them with justification)

---------------------------------------------------------------------------------------------------------------------