---
title: "DATAX500"
author: "1631819"
date: "2026-03-22"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
--------------------------------------------------------------------------------
[View moodle submission](https://elearn.waikato.ac.nz/mod/assign/view.php?id=2241094&action=editsubmission)
--------------------------------------------------------------------------------

# Assignment 1

1. Using the '_heart.csv_' data-set, calculate the following statistical measures
   for age in _R_ **(6 marks)**
     
      a) Mean **(2 marks)**
           
        
        
      b) Median **(2 marks)**
           
        
        
      c) Sample standard deviation **(2 marks)**



----------------------------------------------------------------------------------------------------------------------

2. A researcher wishes to investigate the relationship between gender and 
   cholesterol levels. Using the dataset, perform an appropriate statistical
   test in R and comment on the output. Your answer should include null
   null hypothesis, alternative hypothesis, and p-value **(5 marks)**
   
    - Null hypothesis:
   
    - Alternative hypothesis:
   
    - p-value:
 
----------------------------------------------------------------------------------------------------------------------

3. A researcher wishes to investigate the predictors of maximum heart rate.
   Using the dataset provided, fit an apprpopriate statistical model of your
   choice. Make sure to include age as one of the predictors **(14 marks)**

     a) What are the null and alternative (two-sided) hypotheses associated with age? **(2 marks)**
     
     
     
     
     b) What is the estimated coefficient of age and what does this imply in terms of the relationship between age and        maximum heart rate? **(3 marks)**
     
     
     
     
     c) Comment on the predictors you have included in the model and justify the inclusion of these variables in your         model **(3 marks)**
     
     
     
     
     d) interpret the model output in plain English. Your interpretation must include the p-value associated with each        coefficient (slopes and intercept) test, and R-squared value **(3 marks)**
     
     
     
     
     e) Comment on whether the assumptions of a linear regression model have been met. Include appropriate statistical        plots to support your answer **(3 marks)**
   
 
----------------------------------------------------------------------------------------------------------------------
 
## Data dictionary:

cp: chest pain type

  1. typical angina
  2. atypical angina 
  3. non-anginal pain
  4. asymptomatic

trtbps:

- resting blood pressure (in mm Hg)

chol:

- cholesterol (in mg/dl)

gender: 

1. female;
2. male;

fbs: (fasting blood sugar > 120 mg/dl)

0. False
1. True


rest_ecg: resting electrocardiographic result

0. Normal
1. Having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV)
2. Showing probably or definite left ventricular hypertrophy by Estes' criteria

----------------------------------------------------------------------------------------------------------------------













```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
