---
title: "DATAX500"
author: "1631819"
date: "2026-03-22"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#tinytex::install_tinytex(force = TRUE)
```
--------------------------------------------------------------------------------
[Github Repository](https://github.com/BrewBarred/DATAX500_A1)

[Moodle Submission](https://elearn.waikato.ac.nz/mod/assign/view.php?id=2241094&action=editsubmission)

--------------------------------------------------------------------------------

# Assignment 1

1. Using the '_heart.csv_' data-set, calculate the following statistical 
   measures for age in _R_ **(6 marks)**
   
   
  - First, we load the data-set:
   
      ```{r load, echo=TRUE}
      # load automatically via git files
      heart<-read.csv("Heart.csv")
      
      
      ## Alternatively, load the dataset manually using:
      # heart<-read.csv(file.choose())
      ```
     
  a) Mean **(2 marks)**
           
      ```{r mean, echo=TRUE}      
      # assign age column to "age" var for easier processing
      age<-heart$age
      
      # calculate the mean age
      mean_age<-mean(age)
      mean_age
      
      # round to 2 d.p.
      mean_age<-round(mean_age, 2)
      mean_age
      ```
        
        
  b) Median **(2 marks)**
           
      ```{r median, echo=TRUE}      
      # calculate the median age
      median_age<-median(age)
      median_age
      ```
        
  c) Sample standard deviation **(2 marks)**
  
      ```{r sd, echo=TRUE}      
      # calculate the sample standard deviation
      sd_age<-sd(age)
      sd_age
      
      # round sample standard deviation to 2 d.p.
      sd_age<-round(sd_age, 2)
      sd_age
      ```



----------------------------------------------------------------------------------------------------------------------

2. A researcher wishes to investigate the relationship between gender and 
   cholesterol levels.
   
   Key-points to notice here is that **we are measuring cholesterol levels** 
   based on gender, making **cholesterol** the _dependent variable_, and 
  **gender** the _independent variable_.
   
   Using the dataset, perform an _appropriate statistical test_ in _R_ and 
   interpret the output. Your answer should include a **null hypothesis**, an 
   **alternative hypothesis **, and a  **p-value ** **(5 marks)**
   
   - Since _gender_ (categorical) and _cholesterol_ (continuous) are 
     **independent** from eachother, we conduct an independent (two-sampled) 
     t-test to test the following hypotheses:
   
   
      **Null hypothesis:**
        
        - There is no difference in mean cholesterol levels between genders.
   
      **Alternative hypothesis:**
        
        - There is a significant difference in mean cholesterol levels 
            between genders.
        
      **p-value:**
            
        ```{r p-value, echo=TRUE}
        # peek at the dataset so we can view the column names
        head(heart, n=2)
        
        # perform a t-test on the heart data set between chol and gender
        t.test(chol~gender, data=heart)
        ```
        
      **- Interpretation:**
      
        - The Welch two-sample t-test provides **evidence to reject the null 
          hypothesis**, that **there will be no difference in mean cholesterol 
          levels between genders, _p_=0.003**.
        
        - This interpretation is further supported with **a large t-value of 
          2.98.**
        
        - **Further testing is recommended** to determine the source or 
          magnitude of this cholesterol difference.
      
        
 
----------------------------------------------------------------------------------------------------------------------

3. A researcher wishes to investigate the predictors of maximum heart rate.
   Using the dataset provided, fit an apprpopriate statistical model of your
   choice. Make sure to include age as one of the predictors **(14 marks)**

   Since _maximum heart rate_ and _age_ are both continuous variables, a regression
   model will probably be suitable for this analysis where we test for a
   relationship between ones age and their maximum heart rate.
  
     a) What are the null and alternative (two-sided) hypotheses associated with 
        age? **(2 marks)**
        
        
        **Null hypothesis:** H₀: β₁ = 0
        
          - The regression coefficient for age is zero (i.e., age has no linear 
            effect on maximum heart rate.)
            
            ### Note:
            This means that for every x increase in age, we would not observe an 
            increase in y for heart rate.
        
        
        **Alternative hypothesis:** H₁: β₁ ≠ 0
        
          - There will be a significant relationship between age and maximum
            heart-rate.
            
     
     b) What is the estimated coefficient of age and what does this imply in 
        terms of the relationship between age and maximum heart rate? 
        **(3 marks)**
     
      
     
     
     c) Comment on the predictors you have included in the model and justify the inclusion of these variables in your         model **(3 marks)**
     
     
     
     
     d) interpret the model output in plain English. Your interpretation must include the p-value associated with each        coefficient (slopes and intercept) test, and R-squared value **(3 marks)**
     
     
     
     
     e) Comment on whether the assumptions of a linear regression model have been met. Include appropriate statistical        plots to support your answer **(3 marks)**
   
 
----------------------------------------------------------------------------------------------------------------------
 
## Data dictionary:

cp: chest pain type

  1. typical angina
  2. atypical angina 
  3. non-anginal pain
  4. asymptomatic

trtbps:

- resting blood pressure (in mm Hg)

chol:

- cholesterol (in mg/dl)

gender: 

1. female;
2. male;

fbs: (fasting blood sugar > 120 mg/dl)

0. False
1. True


rest_ecg: resting electrocardiographic result

0. Normal
1. Having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV)
2. Showing probably or definite left ventricular hypertrophy by Estes' criteria

----------------------------------------------------------------------------------------------------------------------













```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
